---
title: "Nonlinear DID: Methods & Theory Extension"
subtitle: "Conditional DID, Nonparametric DID, and Identifiability Beyond Linearity"
author: "Geminos"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    math: mathjax
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(mgcv)
library(MatchIt)
library(patchwork)
set.seed(42)
```

---

# Data Generating Process (Nonlinear)

We generate a panel-style dataset where:

- Treatment assignment $D_i$ follows a **nonlinear propensity score**
- Baseline trends are **nonlinear** in covariates
- Treatment effects are **heterogeneous** across age and genre

## Parameters

| Symbol | Meaning |
|--------|---------|
| $n$ | Sample size = 2000 |
| $X_i = (\text{age}_i, g_i, r_i)$ | Covariates |
| $D_i \in \{0,1\}$ | Treatment (RS indicator) |
| $Y_{it}$ | Outcome at time $t \in \{1,2,3\}$ (pre-pre, pre, post) |
| $\tau_i$ | Individual treatment effect |

## DGP Code

```{r dgp}
n <- 2000

# ── Covariates ──────────────────────────────────────────────
age           <- runif(n, 20, 30)
genre_pref    <- sample(c("Action","Drama","Comedy","SciFi","Romance"), n, replace=TRUE)
major_release <- sample(c("sports","politics","celebrity","tech","news"),   n, replace=TRUE)

genre_num   <- as.numeric(factor(genre_pref))
release_num <- as.numeric(factor(major_release))

# ── Nonlinear propensity score ───────────────────────────────
linpred <- 
  0.5  * sin(age / 3) +
  0.3  * (genre_num^2) / 10 +
 -0.4  * log(release_num + 1) +
  0.2  * (age * genre_num) / 30

ps_true <- plogis(linpred)
D       <- rbinom(n, 1, ps_true)

# ── Pre-period outcomes (nonlinear in X) ─────────────────────
Y_pre1 <- 
   2 * sin(age / 4) +
   0.5 * genre_num^1.5 +
  -0.3 * release_num^2 / 5 +
   rnorm(n, 0, 2) + 30

Y_pre2 <- 
   3 * sin(age / 5) +
   0.7 * genre_num^1.5 +
  -0.4 * release_num^2 / 5 +
   0.5 * age +
   rnorm(n, 0, 2) + 30

# ── Heterogeneous true treatment effect ──────────────────────
tau <- 
  1.5 +
  0.5 * sin(age / 2) +
  0.3 * (genre_num == 1) +   # Action: higher effect
 -0.4 * (genre_num == 2)     # Drama: lower effect

ETT_true <- mean(tau[D == 1])
cat("True ETT (treated units only):", round(ETT_true, 4), "\n")

# ── Post-period outcome ──────────────────────────────────────
Y_post <- 
  Y_pre2 +
  tau * D +
  5 * sin(age / 3) +
  rnorm(n, 0, 2)

# ── Dataset ──────────────────────────────────────────────────
data <- data.frame(
  age, genre_pref, major_release,
  genre_num, release_num,
  D, ps_true,
  Y_pre1, Y_pre2, Y_post,
  diff = Y_post - Y_pre2,
  tau
)
```

## Exploratory Plots

```{r plots, fig.width=10, fig.height=4}
p1 <- ggplot(data, aes(x = age, y = Y_post, color = factor(D))) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("#4477AA","#EE6677"),
                     labels = c("Control","Treated")) +
  labs(title = "Post-outcome vs Age", color = "Group") +
  theme_minimal()

p2 <- ggplot(data, aes(x = age, y = diff, color = factor(D))) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("#4477AA","#EE6677"),
                     labels = c("Control","Treated")) +
  labs(title = "First Difference (Y_post - Y_pre2) vs Age", color = "Group") +
  theme_minimal()

p3 <- ggplot(data, aes(x = age, y = tau)) +
  geom_point(alpha = 0.2, size = 0.8, color = "#228833") +
  geom_smooth(se = FALSE, color = "#228833", linewidth = 1.2) +
  labs(title = "True Individual Treatment Effect τ_i vs Age") +
  theme_minimal()

p1 + p2 + p3
```

---

# DID Estimators

## Unconditional DID (Naive)

Simple two-way difference, ignoring all covariates.

```{r naive-did}
ETT_naive <- 
  (mean(data$Y_post[data$D==1]) - mean(data$Y_pre2[data$D==1])) -
  (mean(data$Y_post[data$D==0]) - mean(data$Y_pre2[data$D==0]))

cat("Unconditional DID:", round(ETT_naive, 4), "\n")
cat("Bias vs true ETT: ", round(ETT_naive - ETT_true, 4), "\n")
```

**Why it's biased:** Treated and control groups differ in their baseline covariates (age, genre, release), and the parallel trends assumption fails marginally because the trends are nonlinear in $X$.

---

## Conditional DID (Misspecified Linear)

We model the counterfactual trend $\mathbb{E}[\Delta Y_i(0) \mid X_i]$ with a linear regression on controls. This is **misspecified** because the true relationship is nonlinear.

```{r linear-did}
ctrl_linear <- lm(
  diff ~ age + genre_pref + major_release,
  data = subset(data, D == 0)
)

# Predict counterfactual trend for treated
cf_linear <- predict(ctrl_linear, newdata = subset(data, D == 1))
ETT_linear <- mean(data$diff[data$D == 1]) - mean(cf_linear)

cat("Linear Conditional DID:", round(ETT_linear, 4), "\n")
cat("Bias vs true ETT:      ", round(ETT_linear - ETT_true, 4), "\n")
```

---

## Conditional DID — GAM (Nonparametric Outcome Regression)

We replace the linear model with a **Generalized Additive Model** to capture nonlinear covariate-outcome relationships.

$$\mathbb{E}[\Delta Y_i(0) \mid X_i] = \alpha + f_1(\text{age}_i) + f_2(g_i) + f_3(r_i)$$

where $f_j$ are estimated smooth functions (cubic regression splines).

```{r gam-did}
ctrl_gam <- gam(
  diff ~ s(age) + genre_pref + major_release,
  data = subset(data, D == 0)
)

cf_gam  <- predict(ctrl_gam, newdata = subset(data, D == 1))
ETT_gam <- mean(data$diff[data$D == 1]) - mean(cf_gam)

cat("GAM Conditional DID:", round(ETT_gam, 4), "\n")
cat("Bias vs true ETT:   ", round(ETT_gam - ETT_true, 4), "\n")
```

---

## Nonparametric DID — IPW (Abadie 2005)

### Propensity Score via GAM

```{r ipw-ps}
ps_gam <- gam(
  D ~ s(age) + genre_pref + major_release,
  family = binomial,
  data = data
)
data$ps_hat <- predict(ps_gam, type = "response")

# Trim propensity scores to avoid extreme weights
trim <- 0.02
data$ps_trim <- pmax(trim, pmin(1 - trim, data$ps_hat))
```

### IPW-DID Estimator (Abadie 2005, eq. 3)

The IPW-DID estimator for the ETT is:

$$\widehat{\text{ETT}}_{\text{IPW}} = \frac{1}{\mathbb{E}[D]} \mathbb{E}\left[\frac{\Delta Y_i \cdot (D_i - \hat{p}(X_i))}{1 - \hat{p}(X_i)}\right]$$

```{r ipw-did}
pD   <- mean(data$D)

ETT_ipw <- mean(
  (data$diff / pD) *
  (data$D - data$ps_trim) /
  (1 - data$ps_trim)
)

cat("IPW-DID (Abadie 2005):", round(ETT_ipw, 4), "\n")
cat("Bias vs true ETT:     ", round(ETT_ipw - ETT_true, 4), "\n")
```

---

## Doubly Robust DID

The doubly robust estimator combines **outcome regression** and **IPW**. It is consistent if *either* the propensity score model or the outcome model is correctly specified.

For the panel ETT, a convenient normalized DR form is:

$$\widehat{\text{ETT}}_{\text{DR}} =
\frac{1}{n}\sum_i \left[
\frac{D_i}{\bar D}
-
\frac{\frac{(1-D_i)\hat p(X_i)}{1-\hat p(X_i)}}{\frac{1}{n}\sum_j \frac{(1-D_j)\hat p(X_j)}{1-\hat p(X_j)}}
\right]
\left(\Delta Y_i - \hat m_0(X_i)\right)$$

where $\hat{m}_0(X_i) = \hat{\mathbb{E}}[\Delta Y_i \mid X_i, D_i=0]$ and $\bar D = n^{-1}\sum_i D_i$.

```{r dr-did}
m0_hat <- predict(ctrl_gam, newdata = data)

phat <- data$ps_trim
pD_bar <- mean(data$D)

# Normalized DR DID for panel ETT
w_treat <- data$D / pD_bar
w_ctrl_raw <- (1 - data$D) * phat / (1 - phat)
w_ctrl <- w_ctrl_raw / mean(w_ctrl_raw)

ETT_dr <- mean((w_treat - w_ctrl) * (data$diff - m0_hat))

cat("Doubly Robust DID:", round(ETT_dr, 4), "\n")
cat("Bias vs true ETT: ", round(ETT_dr - ETT_true, 4), "\n")
```

---

## Kernel Matching DID

We use **Mahalanobis-distance** nearest-neighbor matching on $(age, g, r)$ to find the closest control unit for each treated unit, then take the average difference in first-differences.

```{r kernel-did, message=FALSE, warning=FALSE}
library(MatchIt)

m_out <- matchit(
  D ~ age + genre_num + release_num,
  data = data,
  method = "nearest",
  distance = "mahalanobis",
  ratio = 1
)

matched_data <- match.data(m_out)

ETT_match <- with(matched_data,
  mean(diff[D == 1]) - mean(diff[D == 0])
)

cat("Kernel/Matching DID:", round(ETT_match, 4), "\n")
cat("Bias vs true ETT:   ", round(ETT_match - ETT_true, 4), "\n")
```

---

## Results Summary

```{r summary-table}
results <- tibble(
  Estimator = c(
    "True ETT",
    "Unconditional DID",
    "Linear Conditional DID",
    "GAM Conditional DID",
    "IPW-DID (Abadie 2005)",
    "Doubly Robust DID",
    "Matching DID (Mahalanobis)"
  ),
  Estimate = round(c(ETT_true, ETT_naive, ETT_linear, ETT_gam,
                     ETT_ipw, ETT_dr, ETT_match), 4),
  Bias = round(c(ETT_true, ETT_naive, ETT_linear, ETT_gam,
                 ETT_ipw, ETT_dr, ETT_match) - ETT_true, 4)
)

knitr::kable(results, caption = "ETT Estimates: All Methods vs True Value")
```

```{r plot-results, fig.width=8, fig.height=4}
results %>%
  filter(Estimator != "True ETT") %>%
  mutate(Estimator = fct_reorder(Estimator, abs(Bias))) %>%
  ggplot(aes(x = Estimator, y = Estimate, fill = abs(Bias) < 0.1)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = ETT_true, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("#CC3311","#009988"),
                    labels = c("|Bias| ≥ 0.1", "|Bias| < 0.1")) +
  coord_flip() +
  labs(title = "DID Estimator Comparison",
       subtitle = paste("True ETT =", round(ETT_true,4)),
       y = "ETT Estimate", x = NULL, fill = "") +
  theme_minimal()
```

---

# Scenario Comparison: Why Naive DID Can Look "Okay"

The first DGP is nonlinear, but the marginal violation of parallel trends is still moderate. To make the contrast clearer, we now add a **stress-test DGP** with:

- stronger selection into treatment
- more separation in covariate distributions
- nonlinear interactions in the untreated trend

This second scenario is designed so that unconditional DID has a harder time "accidentally" working well.

```{r scenario-compare}
estimate_did_suite <- function(data) {
  ETT_true <- mean(data$tau[data$D == 1])

  ETT_naive <- 
    (mean(data$Y_post[data$D == 1]) - mean(data$Y_pre2[data$D == 1])) -
    (mean(data$Y_post[data$D == 0]) - mean(data$Y_pre2[data$D == 0]))

  ctrl_linear <- lm(
    diff ~ age + genre_pref + major_release,
    data = subset(data, D == 0)
  )

  cf_linear <- predict(ctrl_linear, newdata = subset(data, D == 1))
  ETT_linear <- mean(data$diff[data$D == 1]) - mean(cf_linear)

  ctrl_gam <- gam(
    diff ~ s(age) + genre_pref + major_release,
    data = subset(data, D == 0)
  )

  cf_gam <- predict(ctrl_gam, newdata = subset(data, D == 1))
  ETT_gam <- mean(data$diff[data$D == 1]) - mean(cf_gam)

  ps_gam <- gam(
    D ~ s(age) + genre_pref + major_release,
    family = binomial,
    data = data
  )

  ps_hat <- predict(ps_gam, type = "response")
  ps_trim <- pmax(0.02, pmin(0.98, ps_hat))
  pD_bar <- mean(data$D)

  ETT_ipw <- mean(
    (data$diff / pD_bar) *
    (data$D - ps_trim) /
    (1 - ps_trim)
  )

  m0_hat <- predict(ctrl_gam, newdata = data)
  w_treat <- data$D / pD_bar
  w_ctrl_raw <- (1 - data$D) * ps_trim / (1 - ps_trim)
  w_ctrl <- w_ctrl_raw / mean(w_ctrl_raw)

  ETT_dr <- mean((w_treat - w_ctrl) * (data$diff - m0_hat))

  m_out <- matchit(
    D ~ age + genre_num + release_num,
    data = data,
    method = "nearest",
    distance = "mahalanobis",
    ratio = 1
  )

  matched_data <- match.data(m_out)
  ETT_match <- with(matched_data,
    mean(diff[D == 1]) - mean(diff[D == 0])
  )

  tibble(
    Estimator = c(
      "True ETT",
      "Unconditional DID",
      "Linear Conditional DID",
      "GAM Conditional DID",
      "IPW-DID (Abadie 2005)",
      "Doubly Robust DID",
      "Matching DID (Mahalanobis)"
    ),
    Estimate = c(
      ETT_true, ETT_naive, ETT_linear, ETT_gam,
      ETT_ipw, ETT_dr, ETT_match
    ),
    Bias = c(
      0, ETT_naive - ETT_true, ETT_linear - ETT_true, ETT_gam - ETT_true,
      ETT_ipw - ETT_true, ETT_dr - ETT_true, ETT_match - ETT_true
    )
  )
}

scenario_a <- results %>%
  mutate(Scenario = "Scenario A: moderate nonlinearity")

n_stress <- 2000

age_s <- runif(n_stress, 20, 30)
genre_pref_s <- sample(c("Action","Drama","Comedy","SciFi","Romance"), n_stress, replace = TRUE)
major_release_s <- sample(c("sports","politics","celebrity","tech","news"), n_stress, replace = TRUE)

genre_num_s <- as.numeric(factor(genre_pref_s))
release_num_s <- as.numeric(factor(major_release_s))

linpred_s <-
  -1.2 +
   1.1 * cos(age_s / 2) +
   0.9 * (genre_pref_s == "Action") +
   0.7 * (genre_pref_s == "SciFi") -
   0.8 * (major_release_s == "politics") +
   1.0 * ((age_s < 24) & (genre_pref_s == "Action")) +
   0.8 * ((age_s > 27) & (major_release_s == "tech"))

ps_true_s <- plogis(linpred_s)
D_s <- rbinom(n_stress, 1, ps_true_s)

Y_pre1_s <-
  1.5 * sin(age_s / 2.5) +
  0.7 * genre_num_s -
  0.4 * release_num_s +
  0.9 * sin(age_s * genre_num_s / 8) +
  rnorm(n_stress, 0, 2) + 28

Y_pre2_s <-
  Y_pre1_s +
  1.2 * cos(age_s / 2.2) +
  0.5 * (genre_pref_s == "Action") -
  0.6 * (major_release_s == "politics") +
  0.8 * sin(age_s * release_num_s / 7) +
  rnorm(n_stress, 0, 1.5)

tau_s <-
  1.4 +
  0.6 * sin(age_s / 1.8) +
  0.5 * (genre_pref_s == "Action") -
  0.4 * (genre_pref_s == "Drama")

m0_stress <-
  2.8 * sin(age_s / 1.7) +
  1.1 * (genre_pref_s == "Action") -
  1.0 * (genre_pref_s == "Drama") +
  1.3 * (major_release_s == "tech") -
  1.1 * (major_release_s == "politics") +
  1.4 * sin(age_s * genre_num_s / 3.5) +
  1.0 * ((age_s > 26) & (major_release_s == "tech"))

Y_post_s <- Y_pre2_s + tau_s * D_s + m0_stress + rnorm(n_stress, 0, 2)

stress_data <- data.frame(
  age = age_s,
  genre_pref = genre_pref_s,
  major_release = major_release_s,
  genre_num = genre_num_s,
  release_num = release_num_s,
  D = D_s,
  ps_true = ps_true_s,
  Y_pre1 = Y_pre1_s,
  Y_pre2 = Y_pre2_s,
  Y_post = Y_post_s,
  diff = Y_post_s - Y_pre2_s,
  tau = tau_s
)

scenario_b <- estimate_did_suite(stress_data) %>%
  mutate(Scenario = "Scenario B: strong selection + interactions")

scenario_compare <- bind_rows(scenario_a, scenario_b) %>%
  mutate(
    Estimate = round(Estimate, 4),
    Bias = round(Bias, 4)
  )

knitr::kable(
  scenario_compare,
  caption = "Two DGPs Compared: Baseline Scenario vs Stress-Test Scenario"
)
```

```{r scenario-compare-plot, fig.width=10, fig.height=5}
scenario_compare %>%
  filter(Estimator != "True ETT") %>%
  ggplot(aes(x = Estimator, y = Bias, fill = Scenario)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  coord_flip() +
  labs(
    title = "Bias Comparison Across Two Nonlinear DID Scenarios",
    subtitle = "Scenario B is constructed to make unconditional DID much less reliable",
    x = NULL,
    y = "Bias"
  ) +
  theme_minimal()
```

**Interpretation:** In Scenario A, unconditional DID can still look "not too bad" because the marginal violation of parallel trends is moderate. In Scenario B, stronger selection and nonlinear interactions generate a larger gap between treated and control counterfactual trends, so naive DID should deteriorate much more visibly.

---

# Theory: Extending Theorem 2.3 to Nonlinear Models

## Setup and Notation

Let $i \in \{1, \ldots, n\}$ index units, $t \in \{0, 1\}$ index time (pre/post). Define:

- $Y_{it}(d)$: potential outcome for unit $i$ at time $t$ under treatment $d \in \{0,1\}$
- $A_i \in \{0,1\}$: binary treatment indicator (received at $t=1$)
- $X_i \in \mathcal{X} \subseteq \mathbb{R}^k$: pre-treatment covariates
- $\Delta Y_i = Y_{i1} - Y_{i0}$: observed first difference
- $\Delta Y_i(0) = Y_{i1}(0) - Y_{i0}(0)$: counterfactual first difference under no treatment

The parameter of interest is the **Effect of Treatment on the Treated (ETT)**:
$$\text{ETT} = \mathbb{E}[Y_{i1}(1) - Y_{i1}(0) \mid A_i = 1]$$

To match the notation used in many theorem statements, this section uses $A_i$ for treatment assignment. The simulation code above keeps the variable name `D`; they refer to the same binary treatment indicator.

---

## Theorem 2.3 (Linear Case — Recall)

In the standard linear DID setup, identification rests on:

> **Assumption L (Linear Parallel Trends):** 
> $\mathbb{E}[\Delta Y_i(0) \mid X_i, A_i] = \mathbb{E}[\Delta Y_i(0) \mid X_i] = X_i^\top \beta$

Under this assumption and no-anticipation, the ETT is identified by:
$$\text{ETT} = \mathbb{E}[\Delta Y_i \mid A_i = 1] - \mathbb{E}[X_i^\top \hat{\beta}_0 \mid A_i = 1]$$
where $\hat{\beta}_0$ is estimated on controls only.

**The linearity constraint $X_i^\top \beta$ is restrictive.** When the true conditional trend is a nonlinear function of $X_i$, this estimator is inconsistent.

---

## Extended Theorem: Nonlinear Identification

### Assumptions

**Assumption 1 (No Anticipation):**
$$Y_{i0}(1) = Y_{i0}(0) \quad \text{a.s.}$$

**Assumption 2 (Nonlinear Conditional Parallel Trends — NCPT):**

There exists a measurable function $\mu_0: \mathcal{X} \to \mathbb{R}$ such that:
$$\mathbb{E}[\Delta Y_i(0) \mid X_i = x, A_i = a] = \mu_0(x) \quad \text{for a.e. } x \in \mathcal{X}, \; a \in \{0,1\}$$

That is, conditional on $X_i$, the **counterfactual trend** is the same for treated and control units — but $\mu_0$ need **not** be linear.

**Assumption 3 (Overlap / Common Support):**
$$0 < p(x) := \mathbb{P}(A_i = 1 \mid X_i = x) < 1 \quad \text{for a.e. } x \in \mathcal{X}$$

**Assumption 4 (Regularity):**
$\mathbb{E}[|\Delta Y_i|^2] < \infty$ and $\mathbb{E}[|\mu_0(X_i)|^2] < \infty$.

---

### Theorem (Nonlinear DID Identifiability)

> **Theorem.** Under Assumptions 1–4, the ETT is identified. Specifically:
>
> $$\text{ETT} = \mathbb{E}[\Delta Y_i \mid A_i = 1] - \mathbb{E}[\mu_0(X_i) \mid A_i = 1]$$
>
> where $\mu_0(x) = \mathbb{E}[\Delta Y_i \mid X_i = x, A_i = 0]$ is identified from the control group.

#### Proof

**Step 1: Decompose the observed first difference for treated units.**

For a treated unit ($A_i = 1$), the observed outcome at $t=1$ is $Y_{i1} = Y_{i1}(1)$. Therefore:
$$\Delta Y_i = Y_{i1}(1) - Y_{i0}(1) \stackrel{A1}{=} Y_{i1}(1) - Y_{i0}(0)$$

Add and subtract $Y_{i1}(0)$:
$$\Delta Y_i = \underbrace{[Y_{i1}(1) - Y_{i1}(0)]}_{\tau_i} + \underbrace{[Y_{i1}(0) - Y_{i0}(0)]}_{\Delta Y_i(0)}$$

**Step 2: Take conditional expectation given $X_i$ within the treated group.**

$$\mathbb{E}[\Delta Y_i \mid X_i = x, A_i = 1] = \mathbb{E}[\tau_i \mid X_i = x, A_i = 1] + \mathbb{E}[\Delta Y_i(0) \mid X_i = x, A_i = 1]$$

By Assumption 2 (NCPT):
$$\mathbb{E}[\Delta Y_i(0) \mid X_i = x, A_i = 1] = \mu_0(x)$$

Therefore:
$$\mathbb{E}[\tau_i \mid X_i = x, A_i = 1] = \mathbb{E}[\Delta Y_i \mid X_i = x, A_i = 1] - \mu_0(x)$$

**Step 3: Integrate over the covariate distribution of the treated group.**

$$\text{ETT} = \mathbb{E}[\tau_i \mid A_i = 1] = \int_{\mathcal{X}} \mathbb{E}[\tau_i \mid X_i = x, A_i = 1] \, d F_{X|A=1}(x)$$

$$= \int_{\mathcal{X}} \left[\mathbb{E}[\Delta Y_i \mid X_i = x, A_i = 1] - \mu_0(x)\right] dF_{X|A=1}(x)$$

$$= \mathbb{E}[\Delta Y_i \mid A_i = 1] - \mathbb{E}[\mu_0(X_i) \mid A_i = 1] \tag{$\star$}$$

**Step 4: Identifiability of $\mu_0$.**

By Assumption 3 (overlap), $\mathbb{P}(A_i = 0 \mid X_i = x) > 0$ a.e. Therefore:
$$\mu_0(x) = \mathbb{E}[\Delta Y_i \mid X_i = x, A_i = 0]$$
is identified from the joint distribution of $(\Delta Y_i, X_i)$ among controls — observed in the data.

**Step 5: All terms in $(\star)$ are identified.**

- $\mathbb{E}[\Delta Y_i \mid A_i = 1]$: observed sample mean of first differences among treated units.
- $\mathbb{E}[\mu_0(X_i) \mid A_i = 1]$: plug in the identified $\mu_0$ evaluated at treated covariate values, then average. This is consistent for any nonparametric estimator of $\mu_0$ satisfying standard regularity conditions.

**Therefore, $\text{ETT}$ is nonparametrically identified.** $\blacksquare$

---

### Corollary: IPW Representation

Under Assumptions 1–4, the ETT also admits the **inverse-probability-weighted** representation:

$$\text{ETT} = \frac{1}{\mathbb{E}[A_i]} \mathbb{E}\left[\frac{(A_i - p(X_i))}{1-p(X_i)} \Delta Y_i\right]$$

**Proof sketch:** By Bayes' rule,
$$dF_{X|A=0}(x) = \frac{1-p(x)}{1-\mathbb{E}[A]} dF_X(x)$$

Substituting into $(\star)$:
$$\mathbb{E}[\mu_0(X_i) \mid A_i=1] = \frac{1}{\mathbb{E}[A]} \mathbb{E}\left[\frac{p(X_i)(1-p(X_i))}{1-p(X_i)} \cdot \frac{\Delta Y_i}{1} \cdot \frac{1-p(X_i)}{p(X_i)} \right] \ldots$$

Expanding the full derivation via iterated expectations recovers the Abadie (2005) IPW formula. $\square$

---

### Corollary: Doubly Robust Representation

Define the doubly robust score:
$$\psi_i = A_i \Delta Y_i - (A_i - p(X_i)) \mu_0(X_i) - \frac{p(X_i)(A_i - p(X_i))}{1-p(X_i)} \Delta Y_i$$

Then $\mathbb{E}[\psi_i] / \mathbb{E}[A_i] = \text{ETT}$ whenever **either** $\mu_0$ or $p$ is correctly specified.

**Key implication for nonlinear settings:** Both $\mu_0$ and $p$ can be estimated nonparametrically (e.g., via GAMs, forests, neural networks). The doubly robust estimator achieves $\sqrt{n}$-consistency even when each nuisance component is estimated at slower rates, provided the product of their errors vanishes fast enough (Chernozhukov et al. 2018, double machine learning).

---

## Comparison: Linear vs Nonlinear Parallel Trends

| Feature | Theorem 2.3 (Linear) | Extended Theorem (Nonlinear) |
|---------|---------------------|------------------------------|
| Trend assumption | $\mu_0(x) = x^\top\beta$ | $\mu_0(x)$ unrestricted |
| Estimation of $\mu_0$ | OLS on controls | Any nonparametric method |
| Identification condition | Linearity + overlap | NCPT + overlap (weaker) |
| Risk of misspecification | High (nonlinear DGP) | Low |
| Rate of convergence of ETT | $\sqrt{n}$ (parametric) | $\sqrt{n}$ (if DR or cross-fitting used) |
| Testable? | Pre-trend F-test | Nonparametric pre-trend test |

---

## Discussion: What NCPT Rules Out

The NCPT assumption is **weaker** than linear parallel trends but still imposes a meaningful restriction:

$$\mathbb{E}[\Delta Y_i(0) \mid X_i, A_i] = \mu_0(X_i)$$

This rules out:

1. **Selection on trends:** Cases where treated units have systematically different counterfactual trends even after conditioning on $X_i$. In this case, the ETT is generally not identified without stronger assumptions (e.g., IV, bounding approaches).

2. **Time-varying treatment effect on control trend:** If $\mu_0$ itself shifts due to spillovers or general equilibrium effects, $\mu_0$ is contaminated.

**Testability via Pre-Trends:**

With two pre-periods ($t=-1, t=0$), NCPT implies:
$$\mathbb{E}[Y_{i0}(0) - Y_{i,-1}(0) \mid X_i, A_i=1] = \mu_0^{\text{pre}}(X_i)$$

which can be tested nonparametrically using the pre-period data (e.g., via a permutation test on $\hat{\mu}_0^{\text{pre}}$ residuals). This is the nonlinear analogue of the parallel pre-trends F-test.

---

# References

- Abadie, A. (2005). Semiparametric difference-in-differences estimators. *Review of Economic Studies*, 72(1), 1–19.
- Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics*, 225(2), 200–230.
- Chernozhukov, V., Chetverikov, D., Demirer, M., et al. (2018). Double/debiased machine learning for treatment and structural parameters. *The Econometrics Journal*, 21(1), C1–C68.
- Sant'Anna, P. H. C., & Zhao, J. (2020). Doubly robust difference-in-differences estimators. *Journal of Econometrics*, 219(1), 101–122.
- Roth, J., Sant'Anna, P. H. C., Bilinski, A., & Poe, J. (2023). What's trending in difference-in-differences? *Journal of Econometrics*, 235(2), 2218–2244.
