---
title: "Report of DAG, Geminos"
output: html_document
date: "Feb 3, 2026"
editor_options: 
  markdown: 
    wrap: 72
---

We summarize below the materials delivered to date for the viewing
behavior simulation project. These items are intended to document our
current modeling assumptions, causal structure, and progress toward a
complete simulated dataset.

```{r include=FALSE, warning=FALSE}
library(dagitty)
library(ggdag)
library(ggplot2)
```

## 1. List of variables and their meaning

The variables in the current data-generating process are grouped into
exogenous variables, predetermined variables, and endogenous outcomes.

```{r include=FALSE, warning=FALSE}
dag_model <- dagitty("
dag {
  age
  genre_preference
  major_release
  rec_syst
  past_minutes_viewed
  subscription
  device_type
  minutes_viewed
  
  device_type -> minutes_viewed
  age -> minutes_viewed
  age -> device_type
  age -> subscription
  
  genre_preference -> minutes_viewed
  genre_preference -> subscription
  
  rec_syst -> minutes_viewed
  
  past_minutes_viewed -> minutes_viewed
  
  subscription -> minutes_viewed
  major_release -> minutes_viewed
}
")

```

### Exogenous variables (not caused by other variables in the model):

-   **Age**: Continuous variable representing the user’s age.

-   **Genre preference**: Categorical variable indicating the user’s
    preferred content genre (Action, Drama, Comedy, SciFi, Romance).

-   **Major release**: Categorical variable capturing large external
    content shocks (e.g., sports events, celebrity news, politics,
    technology, general news).

### Predetermined / sequential variables:

-   **Past minutes viewed**: Continuous variable representing historical
    viewing behavior prior to the current period.

-   **Recommendation system exposure (rec_syst)**: Binary indicator for
    whether the user is exposed to a recommendation system, modeled as a
    function of past viewing behavior.

### Endogenous variables:

-   **Subscription**: Binary variable indicating whether the user holds
    a subscription, influenced by age and genre preference.

-   **Device type**: Categorical variable indicating the primary viewing
    device (TV, Mobile, Tablet, PC), influenced by age.

-   **Minutes viewed**: Continuous outcome variable measuring total
    viewing time in the current period.
-   **alive_i** : A binary indicator of whether the user is still active on the platform.

### Variable for further analysis

-   time_point_0 
The time point that users report past_minutes_viewed

-   time_point_1
The time point that users report minutes_viewed



```{r include=FALSE, warning=FALSE,echo=FALSE}
set.seed(1000)
n <- 1000

data <- data.frame(id = 1:n)

# -------------------------
# Exogenous variables
# -------------------------

# Age
data$age <- round(rnorm(n, mean = 20, sd = 4), 1)

# Genre preference
data$genre_preference <- sample(
  c("Action", "Drama", "Comedy", "SciFi", "Romance"),
  n, replace = TRUE,
  prob = c(0.25, 0.20, 0.30, 0.15, 0.10)
)

# Major release (exogenous content shock)
data$major_release <- sample(
  c("sports", "politics", "celebrity", "tech", "news"),
  n, replace = TRUE,
  prob = c(0.30, 0.20, 0.30, 0.10, 0.10)
)

# -------------------------
# Past behavior (pre-treatment)
# -------------------------

data$past_minutes_viewed <- rnorm(
  n,
  mean = 30 + 0.3 * data$age,
  sd = 8
)

# -------------------------
# Recommendation system
# past behavior → rec system
# -------------------------

data$rec_syst <- rbinom(
  n, 1,
  plogis(-1 + 0.04 * data$past_minutes_viewed)
)

# -------------------------
# Subscription
# age + genre preference → subscription
# -------------------------

data$subscription <- rbinom(
  n, 1,
  plogis(
    -2 +
      0.04 * data$age +
      ifelse(data$genre_preference == "Drama",   0.5, 0) +
      ifelse(data$genre_preference == "SciFi",   0.4, 0) +
      ifelse(data$genre_preference == "Comedy",  0.3, 0) +
      ifelse(data$genre_preference == "Romance", 0.2, 0) +
      ifelse(data$genre_preference == "Action",  0.1, 0)
  )
)

# -------------------------
# Device type
# age → device_type
# -------------------------

data$device_type <- apply(
  plogis(cbind(
    0.6 - 0.03 * data$age,  # TV
    0.2 + 0.04 * data$age,  # Mobile
    0.1,                   # Tablet
    0.1                    # PC
  )),
  1,
  function(p) sample(
    c("TV", "Mobile", "Tablet", "PC"),
    size = 1,
    prob = p / sum(p)
  )
)

# -------------------------
# Outcome: minutes viewed
# -------------------------

data$minutes_viewed <-
  60 +
  0.06 * data$age +
  0.35 * data$past_minutes_viewed +
  6   * data$rec_syst +
  15  * data$subscription +
  ifelse(data$device_type == "TV",     8, 0) +
  ifelse(data$device_type == "Mobile", -4, 0) +
  ifelse(data$major_release == "sports",    10, 0) +
  ifelse(data$major_release == "celebrity",  8, 0) +
  ifelse(data$major_release == "news",      12, 0) +
  ifelse(data$genre_preference == "Action",  4, 0) +
  ifelse(data$genre_preference == "Comedy",  2, 0) +
  rnorm(n, mean = 0, sd = 10)

```

## 2. Proposed design (sample size and time structure)

-   **Number of subjects**: 1,000 users.

-   **Time structure**:

    -   One pre-period capturing historical viewing behavior (past
        minutes viewed).

    -   One outcome period measuring minutes viewed.

-   **Treatment / exposure of interest**:

    -   Recommendation system exposure, which is probabilistically
        assigned based on past viewing behavior.

-   **Purpose of current design**:

    -   To validate the causal structure implied by the DAG.

    -   To demonstrate feasibility of simulating realistic viewing
        behavior under explicit modeling assumptions.

## 3. Directed Acyclic Graph (DAG)

We specify a causal DAG to encode assumed relationships among user
characteristics, platform mechanisms, and viewing outcomes.

```{r echo=FALSE}
p <- ggdag(dag_model, layout = "circle") +
  theme_dag() +
  geom_dag_label(colour = "red")+
  labs(title = "Causal DAG for Viewing Behavior")+
  theme_minimal()
p
```

```{r include=FALSE,echo=FALSE}
fit_true <- lm(
  minutes_viewed ~ 
    age +
    past_minutes_viewed +
    rec_syst +
    subscription +
    device_type +
    major_release +
    genre_preference,
  data = data
)

summary(fit_true)
```

## 4. Conditional Probability Distribution

Each variable in the model is generated according to a conditional
mechanism that reflects how the platform and users interact in practice.

-   Historical viewing behavior is determined first and serves as a
    baseline indicator of user engagement.

-   Recommendation exposure depends on prior engagement, reflecting
    targeting rules commonly used in recommendation systems.

-   Subscription decisions depend on user demographics and content
    preferences.

-   Device choice varies systematically with age.

-   Viewing time in the current period is then generated as the combined
    result of user characteristics, platform interventions, content
    availability, and random variation.
    


## 5. Probability Caculation

P(minutes_viewed,age,past_minutes_viewed,rec_syst,major_release,alive_i,subscription,genre_preference)= 

P(minutes_viewed|age,past_minutes_viewed,rec_syst,major_release,alive_i,genre_preference) 

P(past_minutes_viewed)P(rec_syst)P(major_release)P(alive_i|subscription)

P(subscription|age,genre_preference)P(age)P(genre_preference) 

## 6. Longitudinal DAG

```{r}
dag_longitudinal <- dagitty("
dag {

preference
Age_1
Age_2
Age_3
min_viewed_1
min_viewed_2
min_viewed_3
rec_syst_2

preference -> min_viewed_1
preference -> min_viewed_2
preference -> min_viewed_3

Age_1 -> min_viewed_1
Age_2 -> min_viewed_2
Age_3 -> min_viewed_3

min_viewed_1 -> min_viewed_2
min_viewed_2 -> min_viewed_3


rec_syst_2 -> min_viewed_3

}
")

ggdag(dag_longitudinal, layout = "sugiyama") +
  theme_dag() +
  geom_dag_label(colour = "red") +
  labs(title = "Longitudinal Causal DAG (Board Version)") +
  theme_minimal()
```

```{r}
library(dplyr)

set.seed(2026)

n <- 1000

data <- data.frame(id = 1:n)

data$Age_1 <- round(rnorm(n, 20, 4), 1)
data$Age_2 <- data$Age_1 + 1
data$Age_3 <- data$Age_1 + 2

data$preference <- sample(
  c("Action", "Drama", "Comedy", "SciFi", "Romance"),
  n,
  replace = TRUE,
  prob = c(0.25, 0.20, 0.30, 0.15, 0.10)
)

data$min_viewed_1 <-
  60 +
  0.05 * data$Age_1 +
  rnorm(n, 0, 10)

data$min_viewed_2 <-
  60 +
  0.05 * data$Age_2 +
  0.3  * data$min_viewed_1 +
  rnorm(n, 0, 10)

data$rec_syst_2 <- rbinom(
  n,
  1,
  plogis(-2 + 0.03 * data$Age_2)
)

data$min_viewed_3 <-
  60 +
  0.05 * data$Age_3 +
  0.3  * data$min_viewed_2 +
  5    * data$rec_syst_2 +
  rnorm(n, 0, 10)

final_data <- data %>%
  select(
    id,
    preference,
    Age_1, Age_2, Age_3,
    min_viewed_1,
    min_viewed_2,
    min_viewed_3,
    rec_syst_2
  )

write.csv(final_data,
          "simulated_longitudinal_final.csv",
          row.names = FALSE)
```

